Transfinite Games for Countable Nondeterminism

### 2017-05-08 nondeterminism/nondeterminism.tex
A whole bunch of things.

\newcommand{\stle}{{\;\le_s\;}}
\newcommand{\exle}{\sqsubseteq}
\newcommand{\exlub}{\bigsqcup}
\newcommand{\dv}{{\text{\lightning}}}
\DeclareMathOperator{\pocl}{pocl}
\newcommand{\opto}{\longrightarrow}
\newcommand{\n}{{\mathtt{n}}}
\newcommand{\IfO}{{\mathtt{If0}\;}}
\newcommand{\suc}{{\mathtt{suc\;}}}
\newcommand{\0}{{\mathtt{0}}}
\newcommand{\plot}{\triangleleft}

@@

\subsection{Soundness and adequacy}

@@

>>>>

@@

\subsection{Consistency}

@@

\subsection{The language OBFPCF}

To prove computational adequacy, our first step is to modify our language.  We remove the recursion combinator $\Y_T^\alpha$ and replace it with a set of constants $\Y_T^\alpha$ of type $(T\to T)\to T$, where $\alpha$ ranges over the ordinals.

The operational semantics of the operators $\Y_T^\alpha$ is as follows:
\[
  \begin{tabular}{cc}
    $\inferrule*[right=$\alpha<\beta$]{ }{\Y_T^\beta M \opto M(\Y_T^\alpha M)}$
      & $\inferrule*[right=]{ }{\Y_T^0 M \opto \Omega}$
  \end{tabular}
  \]
Thus, $\Y_T^n$ denotes a recursion that can take at most $n$ steps, $\Y_T^\omega$ denotes a recursion that can take $n$ steps, where $n$ is a number chosen at the start of the evaluation, and so on.  We call this new language \emph{Ordinal Bounded Fair PCF}, or OBFPCF.

Note, however, that it will not be sufficient to define must-convergence for the new language directly from the operational semantics.  Take the factorial function $\nat\to\nat$, for instance, which is usually defined as $\Y F$, where $F=\lambda f.\lambda n.\IfO n (\suc \0) (\lambda m.n\times(f m))$.  The number of recursion steps is unbounded, but is always known at the start of evaluation.  Thus, we should expect to be able to define the factorial function as $\Y_{\nat\to\nat}^\omega F$.  However, if we try to evaluate, say, $1!$, we end up with the following possible evaluation sequence:
\[
  \Y_{\nat\to\nat}^\omega F 1 \opto F (\Y_{\nat\to\nat}^0 F) \;1 \opto \Y_{\nat\to\nat}^0 F\;0\opto\Omega \to \cdots
  \]
which is infinite.  The problem is that we have passed immediately down to $\Y^0$, when we could have completed the evaluation by passing down to $\Y^1$.

\begin{definition}
  Let $M$ be a term of OBFPCF.  We define the \emph{shadow} $\shad M$ of $M$ to be the FPCF term formed by replacing every instance of $\Y_T^\alpha$ with $\Y_T$ for all $\alpha$.

  If $M$ is a term in some programming language, we define an \emph{evaluation tree} to be a rooted tree $T$ with vertices $v$ labelled by terms $T(v)$ of that language such that:
  \begin{itemize}
    \item The root is labelled with $M$
    \item If $v$ is a vertex with label $N$ and $w$ is a child of $v$ with label $P$ then $N\opto P$
    \item Distinct children of the same vertex have distinct labels
  \end{itemize}
  For any term $M$, there is clearly a unique maximal such tree, which we denote by $\tree_M$.

  If $M$ is a term of FPCF, we define a \emph{decoration} of $\tree_M$ to be an evaluation tree $T$ in OBFPCF together with a (depth-preserving) homomorphism of rooted trees $\phi\from T\to\tree_M$ such that for all vertices $v$ in $T$, the label of $\phi(v)$ is the shadow of the label of $v$:
  \[
    \tree_M(\phi(v)) = \shad(T(v))
    \]
  If $M$ is a term of OBFPCF, we say that $M$ \emph{must converge}, and write $M\mustconverge$, if:
  \begin{itemize}
    \item $\shad M\mustconverge$
    \item There is a decoration $T$ of $\tree_{\shad M}$ such that the root of $T$ is labelled with $M$.
  \end{itemize}
\end{definition}

\begin{lemma}
  Let $M$ be a term of FPCF such that $M\mustconverge$.  Then there exists some term $M'$ of OBFPCF such that $M=\shad M'$ and $M'\mustconverge$.  Moreover, for every instance of a constant $\Y_T^\alpha$ in $M'$, we have $\alpha<\omega_1$.
  \begin{proof}
    Since $M\mustconverge$, $\tree_M$ is a well-founded tree.  We need to define a decoration of $\tree_M$.  We do this using the standard definition of the ordinal depth of a tree.  Let $T$ have the same vertices as $\tree_M$.  Then we recursively tag the vertices of $T$ with ordinals as follows: we tag a vertex $v$ with $\alpha$ if $v$ is a leaf of the tree formed by removing all vertices that have been tagged with $\beta<\alpha$.  Since $T$ is well-founded, this process must terminate.  Thus, a vertex $v$ is tagged with the supremum of the successors of the tags of its children.  Since there are at most countably many children of any vertex, we inductively show that the tag on each vertex is the supremum of countably many countable ordinals, so it is countable.

    We then relabel the vertices $v$ of $T$ inductively as follows: first, we go through the tree and decorate every instance of $\Y$ on a vertex $v$ with the tag of $v$.  Presently, this is not a valid evaluation tree, since if a term contains an instance of $\Y$ but does not reduce via the rule
    \[
      \inferrule{ }{\Y_T M \opto M(\Y_T M)}
      \]
    then we are not allowed to decorate the instances of $\Y$ in the child with different ordinals from the instances in the parent.  In this case, redecorate the instances of $\Y$ in the child to match those in the parent.  We continue by induction over the depth of the tree and end up with a valid decoration $\tree_M$.
  \end{proof}
\end{lemma}

\begin{remark}
  One way to think of must-convergence in OBFPCF is as follows: given an OBFPCF term $M$, we define a two-player game as follows: if $M$ reduces to a unique term $N$, then replace $M$ with $N$ and start again.  Otherwise, if $M$ reduces according to the rule:
  \[
    \inferrule*[right=$n$]{ }{\wn\opto\n}
    \]
  then Player I chooses the value of $n$ to replace $\wn$ with and we go back to the beginning.  If $M$ reduces according to the rule
  \[
    \inferrule*[right=$\alpha<\beta$]{ }{\Y_T^\beta M \opto M (\Y_T^\alpha M)}
    \]
  then Player II chooses an ordinal $\alpha<\beta$ and applies the rule, and we start again.  If we converge, then we say that Player II wins.  Otherwise, we say that Player I wins.  We say that $M\mustconverge$ if and only if Player II has a winning strategy for this game.
\end{remark}

\begin{example}
  We illustrate the capabilities of this new language with a small example.  In FPCF, we have a term of type $\nat\to\nat$ given by:
  \[
    \Y_{\nat\to\nat}(\lambda f.\lambda n.\IfO n\;\0\;f)
    \]
  that returns $\0$ for all inputs.

  Given a term $M$ of type $T\to T$ and a natural number $n$, we may inductively define $\Y_T^n M$ up to observational equivalence within PCF as:
  \begin{align*}
    &\Y_T^0 M = \Omega \\
    &\Y_T^{n+1} M = M (\Y_T^n M)
  \end{align*}
  In this case, we see that $\Y_{\nat\to\nat}^n(\lambda f.\lambda n.\IfO n\;\0\;f)$ evaluates to $\0$ if it is passed a number that is less than $n$ and diverges otherwise.  Thus:
  \[
    \Y_{\nat\to\nat}^n(\lambda f.\lambda n.\IfO n\;\0\;f)\;\wn
    \]
  may diverge for all $n$, while $\Y_{\nat\to\nat}(\lambda f.\lambda n.\IfO n\;\0\;f)\;\wn$ converges.  This is reflected in our game semantics: the denotation of $\Y_{\nat\to\nat}^n(\cdots)$ converges for inputs less than $n$ and diverges otherwise; if we pass to the stable least upper bound, then the resulting function converges for all inputs.  Composing the functions with $\wn$ shows us that composition is not continuous with respect to the stable order $\stle$.

  Now consider the following term of OBFPCF:
  \[
    \Y^\omega (\lambda f.\lambda n.\IfO n\;\0\;f)\;\wn
    \]
  A typical evaluation of this term might look like this:
  \begin{align*}
    & \Y^\omega (\lambda f.\lambda n.\IfO n\;\0\;f)\;\wn \\
    \opto\quad & (\lambda f.\lambda n.\IfO n\;\0\;f) \Y^m (\lambda f.\lambda n.\IfO n\;\0\;f)\;\wn \\
    \opto\quad & (\lambda n.\IfO n \;\0 \;\Y^m (\lambda f.\lambda n.\IfO n\;\0\;f))\;\wn \\
    \opto\quad & \IfO \wn\;\0\;\Y^m(\lambda f.\lambda n.\IfO\;n\;\0\;f) \\
    \opto\quad & \IfO (m+1)\;\0\;\Y^m(\lambda f.\lambda n.\IfO n\;\0\;f) \\
    \opto\quad & \Y^m(\lambda f.\lambda n.\IfO n\;\0\;f) \;m
  \end{align*}
  which diverges, as we saw above.  The important point is that whatever value of $m$ we choose to decorate $\Y$ with in the second line, there is a possibility that we will nondeterministically choose a value for $\wn$ that is greater than or equal to $m$ and so causes the program to diverge.  Therefore:
  \[
    \Y^\omega (\lambda f.\lambda n.\IfO n\;\0\;f)\;\wn\not\mustconverge
    \]
  Now, if we repace $\omega$ with $\omega+1$, we arrive at the original behaviour: we can replace the $\omega+1$ with $\omega$ in the first line so we arrive at:
  \begin{align*}
    & \Y^{\omega+1} (\lambda f.\lambda n.\IfO n\;\0\;f)\;\wn \\
    \opto\quad & (\lambda f.\lambda n.\IfO n\;\0\;f) \Y^\omega (\lambda f.\lambda n.\IfO n\;\0\;f)\;\wn \\
    \opto\quad & \cdots \\
    \opto\quad & \Y^\omega(\lambda f.\lambda n.\IfO n\;\0\;f) \;m \\
    \opto\quad & (\lambda f.\lambda n.\IfO n\;\0\;f) \Y^{m+1}(\lambda f.\lambda n.\IfO n\;\0\;f) \;m \\
    \opto\quad & \cdots \\
    \opto\quad & \0
  \end{align*}

Here, we are able to choose which natural number to decay to after the value for $\wn$ has been chosen, so this term must converge.

Note that although $\Y^{\omega+1}(\lambda f.\lambda n.\IfO n\;\0\;f)$ is operationally equivalent to $\Y(\lambda f.\lambda n.\IfO n\;\0\;f)$, $\Y^\omega(\lambda f.\lambda n.\IfO n\;\0\;f)$ has no equivalent in FPCF.  Thus, OBFPCF allows us to discern levels of convergence that are invisible in FPCF.

  Importantly, notice that function application \emph{is} now continuous: $\Y^\omega(\cdots)$ is the limit of the $\Y^n(\cdots)$, and $\Y^\omega(\cdots)\;\wn$ may diverge just as $\Y^n(\cdots)\;\wn$ may.

  At the moment, our game semantics is unable to model terms such as $\Y^\omega(\lambda f.\lambda n.\IfO n\;\0\;f)$.  We will need to modify it.
\end{example}

\subsection{Game semantics for OBFCPF}

\begin{definition}
  Let $A$ be a game.  A \emph{multistrategy} for $A$ is a stably directed set $\Sigma$ of strategies for $A$ such that for all $\sigma,\tau\in \Sigma$ we have $T_\sigma=T_\tau$.

  Given a multistrategy $\Sigma$, we write $T_\Sigma$ for the unique value of $T_\sigma$, where $\sigma\in\Sigma$.

  We compose multistrategies in the obvious way: let $A,B,C$ be games, let $\Sigma$ be a multistrategy for $A\iimpl B$ and let $\Tau$ be a multistrategy for $B\iimpl C$.  Then we define:
  \[
    \icomp\Tau\Sigma = \{\icomp\tau\sigma\suchthat\sigma\in\Sigma,\;\tau\in\Tau\}
    \]
  By Lemmas \ref{compositionMonotonicSO} and \ref{exponentiationMonotonicSO}, this is still stably directed.

  We also extend the stable order to multistrategies.  This time, $\stle$ is a preorder rather than being a full order.

  Let $A$ be a game, and let $\Sigma,\Tau$ be multistrategies for $A$.  We say that $\Sigma\stle\Tau$ if for all $\tau\in\Tau$ there exists $\sigma\in\Sigma$ such that $\sigma\stle\tau$.

  We say that $\Sigma \steq \Tau$ if $\Sigma\stle\Tau$ and $\Tau\stle\Sigma$.
\end{definition}

\begin{lemma}
  Let $A$ be a game, let $\sigma$ be a strategy for $A$ and let $\Sigma$ be a multistrategy for $A$ such that $\{\sigma\}\steq\Sigma$.  Then $\sigma\in\Sigma$ and is a $\stle$-least element of $\Sigma$.
  \begin{proof}
    Since $\Sigma\stle\{\sigma\}$, we know that $\tau\stle\sigma$ for all $\tau\in\Sigma$.  Since $\{\sigma\}\stle\Sigma$, we know that $\sigma\stle\tau$ for some $\tau\in\Tau$ and hence by antisymmetry that $\sigma\in\Sigma$.
  \end{proof}
\end{lemma}

Conversely, if $\Sigma$ is a strategy with a $\stle$-greatest element $\sigma$, then $\{\sigma\}\steq\Sigma$.

\begin{definition}
  Let $A$ be a game and let $F$ be a set of multistrategies for $A$ that is directed with respect to $\stle$.  We define a strategy $\exlub F$ given by
  \[
    \exlub F = \left\{\left(\bigcup_{\Sigma\in F} T_\Sigma, D_\sigma\right)\;\middle|\;\textrm{$\sigma\in\Sigma$ for some $\Sigma\in F$}\right\}
    \]
\end{definition}

\begin{proposition}
  $\exlub F$ is a valid multistrategy for $A$.
  \begin{proof}
    We need to show that for all $\Sigma\in F$ and for all $\sigma\in\Sigma$, $\left(\bigcup_{\Sigma\in F}T_\Sigma,D_\sigma\right)$ satisfies the diverge-or-reply rule.  Let $\Sigma\in F$ and let $\sigma\in\Sigma$.  Suppose that $s\in T_\Tau$ for some $\Tau\in F$, that $sa\in P_A$ and that $sa\not\in D_\sigma$.  Since $F$ is stably directed, there exists some multistrategy $\Upsilon\in F$ such that $\Sigma,\Tau\stle\Upsilon$.  Let $\tau\in\Tau$.  Then there are $\upsilon_1,\upsilon_2\in\Upsilon$ such that $\sigma\stle\upsilon_1$ and $\tau\stle\upsilon_2$.  Then, since $\Upsilon$ is stably directed, there is some $\upsilon\in\Upsilon$ such that $\upsilon_1,\upsilon_2\stle\upsilon$.  In particular, $s\in T_\upsilon$ and $sa\not\in D_\upsilon$.  Therefore, $sab\in T_\upsilon$ for some $b$ and so $sab\in\bigcup_{\Sigma\in F}T_\Sigma$.
  \end{proof}
\end{proposition}

\begin{proposition}
  $\exlub F$ is a $\stle$-least upper bound for $F$.

  \begin{proof}
    Let $\Sigma\in F$ and let $\sigma\in\Sigma$.  Then we have $\sigma\stle\left(\bigcup_{\Sigma\in F}T_\Sigma,D_\sigma\right)\in\exlub F$.  So $\exlub F$ is a $\stle$-upper bound for $F$.

    Now suppose that $\Xi$ is some other $\stle$-upper bound for $F$.  Then for all $\Sigma\in F$ we must have $T_\Sigma\subset T_\Xi$ and so $\bigcup_{\Sigma\in F}T_\Sigma\subset T_\Xi$.  Now suppose that $\left(\bigcup_{\Sigma\in F}T_\Sigma,D_\sigma\right)$ is some element of $\exlub F$, where $\Sigma\in F$ and $\sigma\in\Sigma$.  We have $\Sigma\stle\Xi$, so there exists some $\xi\in\Xi$ such that $\sigma\stle\xi$.  We have $\xi = (T_\Xi, D_\xi)$, where $D_\xi\subset D_\sigma$.  Therefore, $\left(\bigcup_{\Sigma\in F}T_\Sigma,D_\sigma\right)\stle\xi$.  It follows that $\exlub F\stle\Xi$.
  \end{proof}
\end{proposition}

Now let $A$ be a game and let $\Sigma$ be a multistrategy for $A\iimpl A$.  We inductively define an ordinal-indexed sequence of multistrategies $\Sigma^\alpha$ for $A\iimpl A$ as follows:
\begin{itemize}
  \item $\Sigma^0 = \{\bot\}$
  \item $\Sigma^{\alpha+1} = \icomp{\Sigma}{\Sigma^\alpha}$
  \item $\Sigma^{\lambda} = \exlub_{\alpha<\lambda}\Sigma^\alpha$ if $\lambda$ is a limit ordinal
\end{itemize}

This chain is increasing and so, by Hartogs' Lemma, it must stabilize: there will be some $\Sigma^\mu$ such that $\Sigma^\mu\steq\icomp\Sigma{\Sigma^\mu}$; this will be a $\stle$-least fixed point of the map $\Tau\mapsto\icomp\Sigma\Tau$ on $\steq$-equivalence classes of multistrategies for $A$.

Now suppose that $\Sigma = \{\sigma\}$, where $\sigma$ is some strategy for $A$.

$\cdots$
